\singlespacing A closer look at Fixed effects regression in structural equation modeling using \texttt{lavaan}

### 2020-06-19 article.Rmd
Fixed a mistake in the matrix algebra
Fixed another mistake in the algebra
Made major revisions

---
title: \singlespacing A closer look at fixed-effects regression in structural equation modeling using \texttt{lavaan}
author: Henrik Kenneth Andersen
abstract: "This article provides an in-depth look at fixed-effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed-effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/random-effects model. \\par  \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis" 
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
---

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Fixed-effects (FE) regression involves controlling for these time-invariant influences via a number of various methods. It thus drastically reduces the number of potential confounders of the relationship between variables of interest.

@@

# Fixed-effects {#fe}

To begin, let us review the conventional, least squares-based FE model. It is normally written as
\begin{align}
y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it} \label{eq:fe}, 
\end{align}
where $y_{it}$ is the observed outcome variable for unit $i$ at time $t$, $\bm{x_{it}}$ is a $1 \times k$ vector of covariates linked to the outcome by the $k \times 1$ vector of coefficients $\bm{\beta}$. All stable, unit-specific characteristics such as date of birth, country of origin, sex, etc. are captured by the time-constant part of the error $\alpha_{i}$, whereas $\varepsilon_{it}$ is the idiosyncratic error term that varies between units and over time [@Bruederl2015].

At least a subset of all the possible time-invariant, unit-specific characteristics affecting the dependent variable are unobservable. Some, like sex and nationality, will be available in a typical dataset, but others, like (for example) ability or motivation will remain unobserved. If the conditional mean of any of these unit-specific characteristics depends on the model covariates, i.e., $\E[\alpha_{i}|\bm{x_{it}}] \ne \E[\alpha_{i}]$ (which implies $\Cov(\bm{x_{it}}^{\intercal},\alpha_{i}) \ne \bm{0}$), then the coefficient of interest, $\beta$, will be biased.

The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects. We can do so by manipulating Equation \eqref{eq:fe} so that the individual effect, $\alpha_{i}$ is eliminated. In doing so, we no longer need to worry about potential bias due to unit-specific characteristics. This can be achieved in a number of ways (e.g., differencing, least squares dummy variable regression), but the most common method involves subtracting the over-time unit mean from each of the terms: 
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}.
\end{align}
Because the average of something that does not change is that thing itself, the individual effects get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002]. The assumption of strict exogeneity comes from the demeaning, where $\E[\bm{\ddot{x}}_{it}^{\intercal}\ddot{\varepsilon}_{it}] = \E[(\bm{x_{it}} - \bm{\bar{x}_{i}})^{\intercal}(\varepsilon_{it} - \bar{\varepsilon}_{i})] = \E[(\bm{x_{it}} - T^{-1}(\bm{x_{i1}} + \bm{x_{i2}} + ... + \bm{x_{iT}}))(\varepsilon_{it} - T^{-1}(\varepsilon_{i1} + \varepsilon_{i2} + ... + \varepsilon_{iT}))]$ which implies that for $\E[\bm{\ddot{x}_{it}}^{\intercal}\ddot{\varepsilon}_{it}]$ to hold, the covariates must be uncorrelated not just with the contemporaneous errors, but the errors at all timepoints.

## Fixed-effects in structural equation modeling

Analytically, Equation \eqref{eq:fe} applies as well to FE models in SEM as we will see shortly. However, there are a number of differences we must address. First, SEM is a covariance-based approach. This means that the vectors of individual observations are only of interest insomuch as they can be converted into observed covariances. Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ individual vectors of length $N$. Lastly, the notation in SEM differs from traditional least squares-based models. There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
\begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+}, 
\end{align}
where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$. $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables. $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables. $\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables. Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them. That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables. $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$. Lastly $\bm{B}$ is a matrix that holds the regression coefficients.

@@

\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix}
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} 
Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$. By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.

Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet. We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.: 
\begin{align}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
\end{align}
where $\bm{I}$ is the identity matrix. By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
\begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align}
which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].

@@

# Fixed-effects in lavaan {#fesem}

@@

Specifying the most basic fixed-effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components. First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one. I will call the latent variable `a` to stand for $\alpha$:

@@

# An example {#ex1}

@@

Fixed-effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001]. This article provides a focused look at the implementation of the basic model, as well as common extensions using the `lavaan` package in `R`.

@@

The most basic FE-SEM is the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014]. For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of fixed-effects in SEM.

>>>>

---
title: \singlespacing A closer look at Fixed effects regression in structural equation modeling using \texttt{lavaan}
author: Henrik Kenneth Andersen
abstract: "This article provides an in-depth look at Fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of Fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/random-effects model. \\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis" 
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
---

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods. It thus drastically reduces the number of potential confounders of the relationship between variables of interest.

@@

# Panel models {#panel}

To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012]
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i$ at time $t$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$, $\bm{z}_{i}$ is a $M \times 1$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ is the unit-specific effect that represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

We can make stating some of the model assumptions easier by rewriting the model in matrix notation
\begin{align}
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
\end{align}
where $\bm{y}_{i}$ and $\bm{\varepsilon}_{i}$ are $T \times 1$ vectors, $\bm{X}_{i}$ and $\bm{Z}_{i}$ are $T \times K$ and $T \times M$ matrices, respectively, $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones and $\alpha_{i}$ is a scalar. $\bm{\beta}$ and $\bm{\gamma}$ are unchanged from Equation \eqref{eq:gpm}.

We typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019]:

- Linearity: the model is linear in its parameters and $\E[\varepsilon_{it}] = 0$ and $\E[\alpha_{i}] = 0$. 
- Independence: the observations are independent across individuals, but not necessarily across time. 
- Strict exogeneity: $\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] = 0$, i.e., the idiosyncratic error at each point in time is mean independent of $\bm{X}_{i}$ at \textit{all} points in time, $t = 1, ..., T$, as well as $\bm{z}_{i}$ and $\alpha_{i}$ . 
- Error variance: homoscedastic and non-serially correlated errors, i.e., $\Var(\varepsilon_{i}|\bm{X}_{i},\bm{z}_{i}, \alpha_{i}) = \sigma^{2}_{\varepsilon} \bm{I}$.

## Random effects

If we are willing to make some assumptions about the distribution of the individual effects and their relationship with the observed time-varying covariates, we are left with the random effects model [@Wooldridge2002; @Wooldridge2012; @Schmidheiny2019; @Skrondal2004]. Namely, we assume the individual effects are normally distributed, with a constant (homoscedastic) variance conditional on the model covariates, i.e., $\alpha_{i}|\bm{X}_{i}, \bm{z}_{i} \sim N(0, \sigma^{2}_{\alpha})$.^[As long as $\bm{X}_{i}$ contains a constant, the assumption that $\alpha_{i}$ has a mean of zero is unproblematic [@Wooldridge2012].] Further, we assume that the expectation of the individual effects conditional on the model covariates is zero, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] = 0$, which implies they are also uncorrelated.

The random effects model works by defining a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewriting the model in Equation \eqref{eq:gpm} as 
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}.
\end{align}
By adding the individual effects to the composite error in each time period, the composite error becomes serially correlated over time, to give the following conditional covariance matrix for the errors, 
\begin{align}
\Var(\nu | \bm{X}, \bm{z}) & = \bm{\Omega}_{\nu} = 
\begin{pmatrix}
\bm{\Omega}_{\nu 1} & \hdots & 0 & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & \bm{\Omega}_{\nu i} & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & 0 & \hdots & \bm{\Omega}_{\nu N}
\end{pmatrix}
\end{align}
with typical elements 
\begin{align}
\Var(\nu_{i} | \bm{X}_{i}, \bm{z}_{i}) & = \bm{\Omega}_{\nu i} = 
\begin{pmatrix}
\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
\end{pmatrix}
\end{align}
where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$. This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects. Keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components.

## Fixed effects

What if the individual effects are independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$? Then grouping them in with the error will cause $\bm{\beta}$ to be biased [@Wooldridge2002; @Wooldridge2012]. In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach involves \textit{demeaning} the equation [@Bruederl2015]. Demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}
\end{align}
where, for example, $\bar{y}_{i} = T^{-1}\sum_{i = 1}^{T}y_{it}$ (and the other over-time means are calculated analogously), and the variables with the dots above them represent the demeaned versions. 
Because the average of something that does not change is that thing itself, the individual effects, along with any time-invariant predictors, get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002]. The downside to this approach is that no time-invariant predictors can be included in the model.

# Fixed effects in structural equation modeling

Moving from the conventional methods outlined above to SEM, we must state the FE model in a different way. We turn to latent variables to account for time-invariant unobserved heterogeneity. In fact, besides accounting for measurement error and the representation of abstract hypothetical concepts, unobserved heterogeneity is one of the main uses of latent variables in SEM [@Skrondal2004].

We first need to convert the data from stacked, long-format vectors of length $NT$ into $T$ individual vectors of length $N$. To see why this is necessary, consider what effect this has on the vector of responses $y_{it}$. Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$. When we convert the data to wide-format, we get $T$ individual equations,
\begin{align}
y_{1i} & = \alpha_{i} + \varepsilon_{1i} \\
y_{2i} & = \alpha_{i} + \varepsilon_{2i} \\
 & \vdots \\
y_{Ti} & = \alpha_{i} + \varepsilon_{Ti}. \label{eq:wide}
\end{align}
Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$. Otherwise, when $t = s$, the variance is $\Var(\alpha_{i}) + \Var(\varepsilon_{it})$. This is exactly the structure we saw above in a typical element of $\bm{\Omega}_{\nu}$.

And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity. The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}. Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time. It also means that the estimated variance of the latent variable is equal to the average covariance between the dependent variable over time. If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the covariance between all timepoints is constant and thus equal to the average covariance between all timepoints.

To show this, consider the following matrix equation of the nonredundant covariances in a three-wave intercept-only model that just restates Equation \eqref{eq:wide}, and which we can solve easily with least squares:
\begin{align}
\bm{b} & = \bm{A}\bm{x} \\
\begin{pmatrix}
\Var(y_{1}) \\
\Cov(y_{2},y_{1}) \\
\Cov(y_{3},y_{1}) \\
\Var(y_{2}) \\
\Cov(y_{3},y_{2}) \\
\Var(y_{3})
\end{pmatrix} & = 
\begin{blockarray}{ccccc}
 & \psi & \phi_{1} & \phi_{2} & \phi_{3}  \\
 \begin{block}{c(cccc)}
 y_{1}^{2}  & 1 & 1 & 0 & 0  \\
 y_{2}y_{1} & 1 & 0 & 0 & 0  \\
 y_{3}y_{1} & 1 & 0 & 0 & 0  \\ 
 y_{2}^{2}  & 1 & 0 & 1 & 0  \\
 y_{3}y_{2} & 1 & 0 & 0 & 0  \\
 y_{3}^{2}  & 1 & 0 & 0 & 1  \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix}
\end{align}
where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation like so
\begin{align}
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} = 
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix} = 
\begin{pmatrix}
.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
\Var(y_{1}) - \psi \\
\Var(y_{2}) - \psi \\
\Var(y_{3}) - \psi
\end{pmatrix}
\end{align}
which shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time. Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.

Having explained how a latent variable is used to estimate the individual effects, we need to state the FE model using SEM-compatible matrix notation. There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
\begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
\end{align}
where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$. $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables. $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables. $\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables. Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them. That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables. $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$. Lastly $\bm{B}$ is a matrix that holds the regression coefficients.

@@

\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} 
Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$. By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.

Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet. We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.: 
\begin{align}
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
\end{align}
where $\bm{I}$ is the identity matrix. By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
\begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align}
which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].

@@

# Fixed effects in lavaan {#fesem}

@@

Specifying the most basic Fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components. First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one. I will call the latent variable `a` to stand for $\alpha$:

@@

# A simulated example {#ex1}

@@

Fixed effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001]. This article provides a focused look at the implementation of the basic model, as well as common extensions using the `lavaan` package in `R`.

@@

The most basic FE-SEM is the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014]. For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of Fixed effects in SEM.
